---
title: "Bergen Bysykkel 2024: Exploratory and Predictive Analytics"
author: "Syed Amjad Ali"
format:
  html:
    code-fold: true  # Allows code folding
    toc: true        # Adds a table of contents
    toc-location: left  # Optional: places TOC in the left sidebar
editor: visual
---

# Introduction {.tabset .tabset-pills}

We'll use three examples:

1.  Demand for cigarettes (effect of price on quantity demanded)
2.  Demand for Norwegian salmon exports (effect of export price on export demand)
3.  Returns to education (effect of education on wages)

In each example, think about the requirements for a valid instrument. Try to come up with arguments for and against.

## Simple IV {.tabset .tabset-pills}

### Cigarette demand and taxes

```{r load cigarette data}
1+1
```

#### Estimation

We report an OLS regression of consumption on price for reference, the IV estimate using tax as an instrument, and an OLS regression of price on tax.

```{r cigarette simple iv}
2+2
```

#### "Manual" IV estimator

```{r simple iv and ols estimators}
3+3
```

### Salmon export demand and input prices

```{r load salmon data}

```

Look at development in exported quantity and price, and fish meal price.

```{r plot salmon variables}

```

#### Estimation

```{r salmon simple iv}

```

#### Note on $R^2$

### Returns to education and distance

```{r load education data}

```

#### Estimation

```{r education simple iv}

```

## IV with controls {.tabset .tabset-pills}

Try to think about what problem the control(s) solve (if any). What are possible remaining problems (violation of assumptions) in each case?

### Cigarette demand and taxes
